Numerical investigation of viscous flow instabilities through heterogeneous porous media. Advantages and limitations of unstructured and adaptive meshes in 2D and 3D domains under different mobility scenarios.

### 2017-12-13 Papers/ViscousInstabilities/Manuscript_ViscousInstabilities_KC.tex
minor corrections

\cortext[cor1]{Corresponding author.}
\address[UoA]{Mechanics of Fluids, Soils \& Structures Group, School of Engineering, University of Aberdeen, UK}
\address[UFRGS]{Engineering School, Federal University of Rio Grande do Sul, Brazil}
\address[UESC]{Department of Exact and Technological Sciences, State University of Santa Cruz, Bahia, Brazil}

@@

\medskip
%Due to the heterogeneity of geological formations, detailed modelling and simulation of multiphase flows through porous media is crucial for design, control and prediction of the aforementioned processes. 
Description of physics and mechanisms of multiphase flows were reported by \citet{wooding_1976} with focus on capillary pressure and flow regimes. A comprehensive review of force balance at the interface between immiscible fluids and resulting mechanisms for flow instabilities can be found in \citet{homsy_1987}. Flow instabilities can be classified as macroscopic and microscopic, the former is due to imposed boundary conditions, whereas the later is associated to local phenomena at the fluids interface \citep[\eg Kelvin-Helmholtz and Saffman-Taylor instabilities][]{saffman_1959}. This work focuses on flow instabilities in two-phase systems due to viscous and stress forces (often referred as viscous instabilities or \textit{viscous fingering}, Section \ref{section:ViscousInstabilities}).

% \citet{adler_1988} described both microscopic- and macroscopic-scale instabilities in Darcy flows. Microscopic instabilities describe phenomena occurring locally \citep[\eg Kelvin-Helmholtz and Saffman-Taylor instabilities][]{saffman_1959}
% Nevertheless we should refer to the area of flow instabilities  as these, first described by \citet{saffman_1958} \red{(IS THIS TRUE? SAFFMANN MAY HAVE BEEN THE FIRST TO THROUGHLY DISCUSS VISCOUS INSTABILITIES, BUT THERE WERE OTHER SCIENTISTS THAT DESCRIBED FLUID INSTABILITIES BEFORE, E.G., RAYLEIGH(1896), TAYLOR (1955) ETC)}. 
%The first distinction should be made between microscopic and macroscopic instabilities. Microscopic instabilities describe phenomena occurring locally, \eg Kelvin-Helmholtz and Saffman-Taylor instabilities \citep{saffman_1959}, whereas macroscopic instabilities involve two-phase flow systems \citep{bottoni_1992}. This paper focuses on microscopic instabilities in two-phase systems due to viscous and stress forces (often referred as viscous instabilities or \textit{viscous fingering}, Section \ref{section:ViscousInstabilities}).

\medskip
In oil and gas reservoir exploration, viscous and density instabilities are relatively common during water-injection processes. As water and oil interacts, the interface between these fluids moves creating an uneven or fingered flow profile (see Fig.~\ref{fig:simple_case}). Viscous fingering results in inefficient flow sweeping which can bypass significant quantities of recoverable oil and may lead to early breakthrough of water into neighbour production wells. Viscous instabilities are mainly controlled by the mobility ratio (MR) between displacing and the displaced fluids. Other conditions that may also influence the severity of viscous fingering are: heterogeneity (\ie wide spatial porosity and/or permeability distribution), gravitational forces, anisotropic dispersion, non-monotonic viscosity profile etc \citep{budek_2017,nicolaides_2015}.

\medskip 
Multi-fluid flow dynamics in porous media are described by continuity and momentum (extended Darcy's law) conservative equations for each fluid (or phase) with coupling mass terms (\ie density and saturation) appearing in both sets. Advanced numerical methods are in continuous development to accurately represent force balances in Darcean flows. Finite difference methods (FDM) have been extensively used in most industry-standard reservoir simulators~\citep{aziz_1986, chen_2005, chang_1990} with relative success. However, they are often limited to relatively simple geometries representing idealised geological formations through structured quadilateral (2-D) and hexahedral (3-D) grid cells~\citep{mlacnik_2004, king_1999}. Additionally, FDM produce excessive numerical dispersion when strong heterogeneity (represented by permeability and porosity fields) is present~\citep{chavent_1986}.

\medskip
The geometrical flexibility associated with high-order numerical accuracy of finite element methods (FEM) has proven to be more efficient than FDM to solve fluid flow dynamics in complex geometries. %FEM discretisation is based upon piecewise representation of the solution space and the computational domain is divided into smaller subdomains (elements) where the solution in each element is constructed from basis functions. %Advantages of FEM over FVM are the better description of the geometry and complex shapes, better treatment of fluid flow and more accurate field-solution calculations using a smaller number of nodes, less memory space and shorter run times. %At this point let us remind that a comparison of computaional methods is difficult and depends always on the test-case.
%The geometrical flexibility associated with high-order numerical accuracy of finite element methods (FEM) has proven to be more efficient compared with FDM schemes when they are used to solve complex geometrical domains. FEM discretization is based upon a piecewise representation of the solution in terms of specified basis functions. The computational domain is divided into smaller domains (elements) and the solution in each element is constructed from the basis functions. Advantages of FEM over FVM are the better description of the geometry and complex shapes, better treatment of fluid flow and more accurate field-solution calculations using a smaller number of nodes, less memory space and shorter run times. 
Among FEM-based formulations for porous media, the control volume finite element methods~\citep[CVFEM,][]{fung_1992} has been widely used as it can guarantee local mass conservation and high-order numerical accuracy as well as being able to use tetrahedral geometry-conforming elements. In traditional CVFEM formulations, pressure and velocity are interpolated using piecewise linear FE basis functions, %$\underline{\bf v}$ and $\underline{\bf p}$)
while material properties and flow conditions (\eg phase saturation, density, temperature, species concentration etc) are represented with CV basis functions~\citep{voller_2009}. Saturation equations are solved explicitly after solving for the dual pressure-velocity at each non-linear iteration~\citep[a detailed description of the implicit pressure explicit saturation, IMPES, formulation can be found in][]{chen_2006,lux_2012}.

@@

\medskip
Since geometries are captured by finite elements, constructed control volumes typically extend on each side of the interface which may have different properties. Therefore, some average values of the coupled velocity-pressure fields (defined in the FE space but projected onto the CV space) are applied across the CVs at these interfaces. These often lead to excessive numerical dispersion especially in highly heterogeneous media (represented by spatial-dependent permeability and porosity fields). In order to overcome such artificial numerical dispersion, a discontinuous hybrid finite element finite volume method (DFEFVM) formulation was introduced by \citet{nick_2011b, nick_2011a}. This novel discretisation scheme was designed to simulate flows through discrete fractured rocks in which CVs are divided along the interfaces of different materials. \citet{cumming_2011} demonstrated that CVFEM discretisation could also be used to solve Richards' equations (coupled mass conservation and Darcy equation) in heterogeneous media with relatively small computational overhead~\citep[compared with traditional coupled velocity-pressure based formulations, see also][]{cumming_phd2012}. Fluxes over CVs were calculated based upon material properties, whereas the saturation field was volume-averaged at the interface of the materials, enforcing mass balance as described by \citet{kirkland_1992}.

@@

\medskip 
In this work, a novel CVFEM formulation, previously introduced by \citet{gomes_2017}\citep[see also][]{jackson_2015,salinas2015}, is used to conduct numerical investigation of formation and evolution of viscous fingers in heterogeneous media. The continuity equation is embedded into the pressure equation to enforce mass conservation whilst ensuring that the force balance is preserved. A hybrid family of \PN[n]{m} triangular and tetrahedral FE pairs is used to discretise velocity and pressure.%, in which velocity is represented by $n^{\text{th}}$-order polynomials that are discontinuous across elements whereas pressure is represented by $m^{\text{th}}$-order polynomials that are continuous across the elements. 
A sketch of the \PN[1]{2} FE-pair is shown in Fig.~\ref{fig:fem_cv}, in which velocity is represented by discontinuous and piecewise linear basis functions whereas pressure is interpolated through continuous and piecewise quadratic basis functions. Scalar fields are stored in CV space (Fig.~\ref{fig:fem_elem}). The dual pressure and velocity fields are represented simultaneously (through non-linear projections) in FE and CV spaces.

@@

The two-phase immiscible and incompressible fluid flow through a porous media domain $\Omega$, may be described by the coupled extended Darcy and saturation equations,

\begin{eqnarray}
\left(\displaystyle\frac{\mu_{\alpha}S_{\alpha}}{{\mathbf K}\mathcal{K}_{r\alpha}}\right) {\mathbf u}_{\alpha} = \underline{\underline{\sigma}}_{\alpha} {\mathbf u}_{\alpha} = -\nabla p + \mathcal{S}_{u,\alpha} \hspace{1cm} \text{ with } x_{i}\in\Omega, t>0, \label{eqn:darcy_eqn} \\
\phi\displaystyle\frac{\partial S_{\alpha} }{\partial t} +   \nabla \cdot \left( {\mathbf u}_{\alpha}  S_{\alpha}\right) =  \mathcal{S}_{cty,\alpha} \hspace{1cm} \text{ with } x_{i}\in\Omega, t>0,\label{eqn:saturation_eqn}
\end{eqnarray}

\noindent respectively, where $\mu$, ${\bf K}$, $p$ and $\phi$ are viscosity, absolute permeability, pressure and porosity, respectively. ${\mathbf u}_{\alpha}$ is the saturation-weighted Darcy velocity of the $\alpha$-phase and $\mathcal{K}_{r,\alpha}$ is the relative permeability. $\mathcal{S}$ is the source term associated with the Darcy and continuity equations. $S_{\alpha}$ represents the saturation of the $\alpha$-phase with mass conservation constraints of 
\begin{displaymath}
\sum\limits_{\alpha=1}^{\mathcal{N}_{p}} S_{\alpha} = 1, 
\end{displaymath}
where $\mathcal{N}_{p}$ denotes the number of phases. $\underline{\underline{\sigma}}_{\alpha}$ is an absorption-like term that represents the implicit linearisation of the viscous frictional forces, and lies in both control volume (CV) and finite element (FE) spaces.

@@

%\end{comment}
%%%%%%%%%%%%%%%
\medskip
The formulation introduced here is applicable to $\mathcal{N}_{p}$ fluid phases and is based on two families of FE-pairs: \PNDG[n]{m} and \PN[n]{m}~\citep{cotter_2009a}, consistent with the dual pressure-velocity representation in CV space. In these families of FE-pairs, velocity is represented by $n^{\text{th}}$-order polynomials that are discontinuous across elements, whereas pressure is represented by $m^{\text{th}}$-order polynomials that may be either continuous or discontinuous (thus the notation \PN[n]{m} and \PNDG[n]{m}, respectively) across elements. Mass balance (continuity) equations are solved in CV space and a Petrov-Galerkin FEM is used to obtain high-order fluxes on CV boundaries, which are limited to yield bounded fields (\eg positive densities, saturations bounded between 0 and 1, etc). Simulations performed for this work were conducted using two types of elements: \PN[1]{1} and \PN[1]{2}.

\medskip
Finite element basis functions for velocity and pressure fields are introduced in the discretisation of force-balance equations (Eqn.~\ref{eqn:darcy_eqn}). Hybrid basis functions are also used to allow CV-based velocity to be extrapolated across the entire element. The extended Darcy equation (Eqn.~\ref{eqn:darcy_eqn}) is discretrised using a FE representation of $\mathbf{v}_{\alpha}=\underline{\underline{\sigma}}_{\alpha}\mathbf{u}_{\alpha}$ and $p$ with FE basis functions $Q_{j}$ and $P_{j}$, respectively. Each component of the weak form of the extended Darcy equation (Eqn.~\ref{eqn:darcy_eqn}) is tested with the $\mathbf{v}_\alpha$ basis function, $Q_{j}$, to obtain:
\begin{eqnarray}
  \sum\limits_{E} \left. \int\limits_{\Omega_E} { {Q}}_{i} \left({\mathbf v}_\alpha + \nabla p  -{\mathbf s}_{u_\alpha} \right) dV \right. + \displaystyle \oint_{\Gamma_{E}} {Q}_i {\mathbf n} \left(p - \tilde{p}\right) d\Gamma + \nonumber \\ 
  \oint_{\Gamma_{\Omega}} {Q}_i {\mathbf n} \left(p - p_\text{bc}\right) d\Gamma = \bm{0},
                   \label{force-semi-disc} 
\end{eqnarray} 
where $\Omega_E$ and $\Gamma_{E}$ are the volume and boundary of element $E$, respectively, and $\Gamma_{\Omega}$ is the boundary of the computational domain. The numerical pressure $\tilde{p}$ appearing in the jump condition (second term in Eqn.~\ref{force-semi-disc}) is the arithmetic mean of the potentially discontinuous pressure across the element $E$. This term vanishes when a continuous formulation is used to discretise the pressure field. The last term in Eqn.~\ref{force-semi-disc} is used to weakly enforce the pressure level to $p_\text{bc}$ on a computational domain boundary.

%\begin{displaymath}
%   \underline{\underline{\sigma}}_{\alpha} = \displaystyle\frac{\mu_{\alpha}S_{\alpha}}{{\mathbf K}\mathcal{K}_{r\alpha}}.
%\end{displaymath}

\medskip
Whilst saturation (and all saturation-dependent material properties such as relative permeability and capillary pressure) is calculated in CV space, absolute permeability is assumed piecewise constant in FE space. Saturation equations (Eqn.~\ref{eqn:saturation_eqn}) are discretised in space with CV basis function, $M_{i}$, and with the $\theta$-method in time \citep{gomes_book_2012}. Velocities across CV interfaces (within and between elements) are calculated through a directional-weighted flux-limited scheme based on upwind value of $\sigma$ at individual CV as described by \citet{jackson_2013}. Summing the discretised Eqn.~\ref{eqn:saturation_eqn} over all phases yields the global mass balance equation,
\begin{eqnarray}
 && \sum_{\alpha=1}^{N_{p}} \left\{\int\limits_{\Omega_{CVi}} \; M_{i} \; \frac{\phi\left(S_{\alpha i}^{n+1}-S_{\alpha i}^{n}\right)}{\Delta t} dV\right.  + \nonumber\\
 &&  \oint_{\Gamma_{CVi}} \left[\; \theta^{n+1/2}\; {\mathbf n}\cdot {\mathbf u}_{\alpha}^{n+1}S_{\alpha}^{n+1} \; + \; \left(1-\theta^{n+1/2}\right) \; {\mathbf u}_{\alpha}^{n}S_{\alpha}^{n}\right] d\Gamma \;- \nonumber\\
 &&  \left.\int_{\Omega_{CVi}} M_{i} \; \mathcal{S}_{cty,\alpha}^{n+\theta} \; dV\;\right\} =0,
\label{global_mass_balance}
\end{eqnarray}
%\medskip %Saturation, and all saturation-dependent material properties such as relative permeability and capillary pressure, are represented in CV space. %Hence, each CV (and FE) is associated with a unique set of petrophysical properties, in contrast to most previous methods in which CVs span multiple values of petrophysical properties described FE-wise.

@@

%\begin{equation}
%\sum  S_{p}-1
%\label{saturation_constrain}
%\end{equation}
\noindent
where $\Omega_{CVi}$ and $\Gamma_{CVi}$ are the volume and boundary of CV i respectively, \textbf{n} is the outward pointing unit normal vector to the surface of $CV_{i}$ and $n$ is the current time level. $\theta$ varies smoothly between $0.5$ (corresponding to Crank-Nicolson method) and $1$ (corresponding to backward-Euler scheme) to avoid the introduction of spurious oscillations for large grid Courant numbers. The discretised global mass and force balance equations (Eqns.~\ref{force-semi-disc}-\ref{global_mass_balance}) are solved using a multigrid-like approach as described by \citet{pavlidis2016} \citep[see also][]{salinas_2016}. The numerical formulation is fully described by \citet{gomes_2017} \citep[see also][]{salinas2015,adam_2016}.

@@

\medskip
As mentioned above viscous flow instabilities can be found across many research areas, disciplines and scales, from chemical and medical separation systems to large scale geological reservoir fields. \citet{muskat_1934} investigated the flow of a fluid injected between two infinite parallel plates (Hele-Shaw cell) and the impact on capillary number ($N_{c}$),

\begin{equation}
N_{c} = \frac{\mu U}{\gamma}.
\label{eq:capillary_number}
\end{equation}

\noindent In Eqn.~\ref{eq:capillary_number}, $\mu$ is the viscosity of the fuid that initially fills the Hele-Shaw cell, $\gamma$ is the surface tension and $U$ is a velocity characteristic of the moving interface. In the Hele-Shaw cell, a fluid is introduced between two parallel flat plates separated by an infinitesimally small gap. This experimental apparatus enables instability to be qualitatively verified and simplifies the problem of flow in both porous and non-porous media.   
%The governing equation of Hele-Shaw flows is identical to that of the inviscid potential flow and to the flow of fluid through a porous medium (Darcy's law) since the Hele-Shaw apparatus may be used for both porous and non-porous media.  
More recently, \citet{howison_2000} and \citet{praud_2005} gave a comprehensive description of the mathematical formulation of immiscible two-phase flows in Hele-Shaw cells (also know as Saffman $\&$ Taylor problem). For a Hele-Shaw cell of a given size the flow development only depends on the capillary number. Thus if \textit{$N_{c}$} is too high \citet{saffman_1959}, \citet{homsy_1987}, \citet{tabeling_1987} and others agreed that the flow develops into a single steady-state finger which moves through the cell with constant velocity $U$.

@@

\medskip 
In heterogeneous domains, such instability may also be triggered by permeability differences across regions as demonstrated in Section~\ref{section:results_homo_hete}. The higher the velocity of the low viscosity fluid, the less wide (tip-splitting behaviour) the finger is. Pressure differences acting on the interface produces a net pressure force,

\begin{equation} 
\Delta P= - \gamma \nabla\cdot\hat{n}. 
\label{eq:pressure_dif} 
\end{equation}

@@

\medskip
As demonstrated by \citet{habermann_1960} \citep[see also][]{budek_2017}, mobility ratio (MR) is a key-parameter to assess fluid displacement and is defined as the ratio of mobility of the displacing (fluid $i$) to that of the displaced fluid ($j$),
\begin{equation}
 \text{MR} \; = \frac{\mathcal{K}_{ri} \; \mu_{j}}{\mathcal{K}_{rj} \; \mu_{i}}. 
\label{eq:MR}
\end{equation}

\noindent MR is a function not only of fluids' viscosity, but also of the parameterised relative permeability, $\mathcal{K}_{r\alpha}$, which is often expressed as a function of local, residual and maximum phase saturations prescribed in the pore rock matrix. In the simulations conducted for this work, the modified Brooks-Corey model was used \citep{alpak_1999,brooks_1964},

\begin{eqnarray}
  \mathcal{K}_{rw}\left(S_{w}\right) &=& \mathcal{K}^{\circ}_{rw}\left[\frc{S_{w}-S_{w,irr}}{1-S_{w,irr}-S_{nw,r}}\right]^{n_{w}}, \label{Eqn:CoreyBrooks1}\\
  \mathcal{K}_{rnw}\left(S_{nw}\right) &=& \mathcal{K}^{\circ}_{rnw}\left[\frc{S_{nw}-S_{nw,r}}{1-S_{w,irr}-S_{nw,r}}\right]^{n_{nw}}, \label{Eqn:CoreyBrooks2}
\end{eqnarray}

@@

\medskip
Here, 2D simulations were conducted in a 5$\times$5 cm domain (Fig.~\ref{fig:homoheleshaw_VN3}a) fully saturated with a fluid. Wetting phase fluid is driven from the bottom left-hand corner of the domain with velocity of 1 cm.s$^{-1}$ (magnitude). Newmann (\ie no flux) boundary conditions were imposed to all borders of the domain except at the top right-hand side corner (named as sink), and a pressure gradient between source and sink regions was initially imposed to the system.

>>>>

\cortext[cor1]{Corresponding author.}
\address[UoA]{Fluids \& Structures Group, School of Engineering, University of Aberdeen, UK}
\address[UFRGS]{Engineering School, Federal University of Rio Grande do Sul, Brazil}
\address[UESC]{Department of Exact and Technological Sciences, State University of Santa Cruz, Bahia, Brazil}

@@

\medskip
%Due to the heterogeneity of geological formations, detailed modelling and simulation of multiphase flows through porous media is crucial for design, control and prediction of the aforementioned processes. 
Description of physics and mechanisms of multiphase flows were reported by \citet{wooding_1976} with focus on capillary pressure and flow regimes. A comprehensive review of mechanisms of fluid instabilities (\ie fingering) and forces acting on the interface between two immiscible fluids can be found in \citet{homsy_1987}. \citet{adler_1988} described both microscopic and macroscopic scale Darcy-equations.
% Nevertheless we should refer to the area of flow instabilities  as these, first described by \citet{saffman_1958} \red{(IS THIS TRUE? SAFFMANN MAY HAVE BEEN THE FIRST TO THROUGHLY DISCUSS VISCOUS INSTABILITIES, BUT THERE WERE OTHER SCIENTISTS THAT DESCRIBED FLUID INSTABILITIES BEFORE, E.G., RAYLEIGH(1896), TAYLOR (1955) ETC)}. 
The first distinction should be made between microscopic and macroscopic instabilities. Microscopic instabilities describe phenomena occurring locally, for example, Kelvin-Helmholtz and Saffman-Taylor instabilities \citep{saffman_1959}. On the other hand, macroscopic instabilities involve two-phase flow system \citep{bottoni_1992}. This paper focuses on microscopic instabilities in two-phase systems that occurs due to viscous and stress forces, often referred as viscous instabilities or \textit{viscous fingering} (see section \ref{section:ViscousInstabilities}).

\medskip
In oil and gas reservoir exploration, viscous and density instabilities are relatively common during water-injection processes. As water and oil interacts, the interface between these two fluids moves creating an uneven or fingered flow profile (see Fig.~\ref{fig:simple_case}). Viscous fingering results in inefficient flow sweeping which can bypass significant quantities of recoverable oil and, may lead to early breakthrough of water into neighbour production wells. Viscous instabilities are mainly governed by the mobility ratio (MR) between the displacing and the displaced fluids. Other conditions that may also influence the severity of viscous fingering are: heterogeneity (\ie wide spatial permeability distribution), gravitational forces, anisotropic dispersion, non-monotonic viscosity profile etc.

\medskip 
Conservation of mass and Darcy's law are the governing equations that describe flow dynamics in porous media. Advanced numerical methods are in continuous development to accurately represent force balances in Darcean flows. Finite difference methods (FDM) have been extensively used in most industry-standard reservoir simulators~\citep{aziz_1986, chen_2005, chang_1990} with relative success. However, they are often limited to relatively simple geometries representing idealised geological formations through structured quadilateral (2-D) and hexahedral (3-D) grid cells of low quality~\citep{mlacnik_2004, king_1999}. Additionally, FDM produce excessive numerical dispersion when heterogeneity (represented by permeability and porosity fields) is present~\citep{chavent_1986}.

\medskip
The geometrical flexibility associated with high-order numerical accuracy of finite element methods (FEM) has proven to be more efficient than FDM schemes when they are used to solve complex geometries. %FEM discretisation is based upon piecewise representation of the solution space and the computational domain is divided into smaller subdomains (elements) where the solution in each element is constructed from basis functions. %Advantages of FEM over FVM are the better description of the geometry and complex shapes, better treatment of fluid flow and more accurate field-solution calculations using a smaller number of nodes, less memory space and shorter run times. %At this point let us remind that a comparison of computaional methods is difficult and depends always on the test-case.
%The geometrical flexibility associated with high-order numerical accuracy of finite element methods (FEM) has proven to be more efficient compared with FDM schemes when they are used to solve complex geometrical domains. FEM discretization is based upon a piecewise representation of the solution in terms of specified basis functions. The computational domain is divided into smaller domains (elements) and the solution in each element is constructed from the basis functions. Advantages of FEM over FVM are the better description of the geometry and complex shapes, better treatment of fluid flow and more accurate field-solution calculations using a smaller number of nodes, less memory space and shorter run times. 
Among FEM-based formulations for porous media, the control volume finite element methods~\citep[CVFEM,][]{fung_1992} has been largely used as it can guarantee local mass conservation and high-order numerical accuracy as well as being able to use tetrahedral geometry-conforming elements. In traditional CVFEM formulations, pressure and velocity are interpolated using piecewise linear FE basis functions, %$\underline{\bf v}$ and $\underline{\bf p}$)
while material properties and flow conditions (\eg phase saturation, density, temperature, species concentration etc) are represented with CV basis functions~\citep{voller_2009}. Saturation equations are solved explicitly after solving for the dual pressure-velocity at each non-linear iteration~\citep[a detailed description of the implicit pressure explicit saturation, IMPES, formulation can be found in][]{chen_2006,lux_2012}.

@@

\medskip
Since geometries are captured by finite elements, constructed control volumes typically extend on each side of the interface which may have different properties. Therefore, some average values of the coupled velocity-pressure fields (defined in the FE space but projected into the CV space) are applied across the CVs at these interfaces. These often lead to excessive numerical dispersion especially in highly heterogeneous media (represented by spatial-dependent permeability and porosity fields). In order to overcome such artificial numerical dispersion, a discontinuous hybrid finite element finite volume method (DFEFVM) was introduced by \citet{nick_2011b, nick_2011a}. This novel discretisation scheme was designed to simulate flows through discrete fractured rocks in which CVs are divided along the interfaces of different materials. \citet{cumming_2011} demonstrated that CVFEM discretisation could also be used to solve Richards' equations (coupled mass conservation and Darcy equation) in heterogeneous media with relatively small computational overhead~\citep[compared with traditional coupled velocity-pressure based formulations, see also][]{cumming_phd2012}. Fluxes over CVs were calculated based upon material properties, whereas the saturation field was volume-averaged at the interface of the materials, enforcing mass balance as described by \citet{kirkland_1992}.

@@

\medskip 
In this work, a novel CVFEM formulation, previously introduced by \citet{gomes_2017}\citep[see also][]{jackson_2015,salinas2015}, is used to conduct numerical investigation of formation and evolution of viscous fingers in heterogeneous media. The continuity equation is embedded into the pressure equation to enforce mass conservation whilst ensuring force balance is preserved. A hybrid family of \PN[n]{m} triangular and tetrahedral FE pairs is used to discretise velocity and pressure, in which velocity is represented by $n$ th-order polynomials that are discontinuous across elements whereas pressure is represented by $m$ th-order polynomials that are continuous across the elements. A sketch of a \PN[1]{2} FE-pair is shown in Fig.~\ref{fig:fem_cv}, in which velocity is represented by discontinuous and piecewise linear basis functions whereas pressure is interpolated through continuous and piecewise quadratic basis functions. Scalar fields are stored in CV space (Fig.~\ref{fig:fem_elem}). The dual pressure and velocity fields are represented simultaneously (through non-linear projections) in FE and CV spaces.

@@

\medskip
The two-phase immiscible and incompressible fluid flow through a porous media domain $\Omega$, may be described by the coupled extended Darcy and saturation equations,

\begin{eqnarray}
\left(\displaystyle\frac{\mu_{\alpha}S_{\alpha}}{{\mathbf K}\mathcal{K}_{r\alpha}}\right) {\mathbf u}_{\alpha} = \underline{\underline{\sigma}}_{\alpha} {\mathbf u}_{\alpha} = -\nabla p + \mathcal{S}_{u,\alpha},\;\;\; x_{i}\in\Omega, t>0 \label{eqn:darcy_eqn} \\
\phi\displaystyle\frac{\partial S_{\alpha} }{\partial t} +   \nabla \cdot \left( {\mathbf u}_{\alpha}  S_{\alpha}\right) =  \mathcal{S}_{cty,\alpha},\;\;\; x_{i}\in\Omega, t>0\label{eqn:saturation_eqn}
\end{eqnarray}

\noindent respectively, where $\mu$, ${\bf K}$, $p$ and $\phi$ are viscosity, absolute permeability, pressure and porosity, respectively. ${\mathbf u}_{\alpha}$ is the saturation-weighted Darcy velocity of the $\alpha$-phase and $\mathcal{K}_{r,\alpha}$ is the relative permeability. $\mathcal{S}$ is the source term associated with the Darcy and continuity equations. $S_{\alpha}$ represents the saturation of the $\alpha$-phase with mass conservation constraints of 
\begin{displaymath}
\sum\limits_{\alpha=1}^{\mathcal{N}_{p}} S_{\alpha} = 1, 
\end{displaymath}
where $\mathcal{N}_{p}$ denotes the number of phases.

@@

%\end{comment}
%%%%%%%%%%%%%%%

\medskip
Finite element basis functions for velocity and pressure fields are introduced in the discretisation of force-balance equation (Eqn.~\ref{eqn:darcy_eqn}). Hybrid basis functions are also used to allow CV-based velocity to be extrapolated across the entire element. $\underline{\underline{\sigma}}_{\alpha}$ is an absorption-like term that represents the implicit linearisation of the viscous frictional forces, and lies in both CV and FEM spaces.
%\begin{displaymath}
%   \underline{\underline{\sigma}}_{\alpha} = \displaystyle\frac{\mu_{\alpha}S_{\alpha}}{{\mathbf K}\mathcal{K}_{r\alpha}}.
%\end{displaymath}

\medskip
Whilst saturation (and all saturation-dependent material properties such as relative permeability and capillary pressure) is calculated in CV space, permeability is piecewise constant in FE space. Saturation equations (Eqn.~\ref{eqn:saturation_eqn}) are discretised in space with CV basis function and with the $\theta$-method in time as described by \citet{gomes_book_2012}. Velocities across CV interfaces (within and between elements) are calculated through a directional-weighted flux-limited scheme based on upwind value of $\sigma$ at individual CV as described by \citet{jackson_2013}.

\medskip
This formulation is applicable to $\mathcal{N}_{p}$ fluid phases and is based on two families of  FE-pairs: \PNDG[n]{m} and \PN[n]{m}~\citep{cotter_2009a}, consistent with the dual pressure-velocity representation in control volume (CV). In these families of FE-pairs velocity is represented by $n^{\text{th}}$-order polynomials that are discontinuous across elements, whereas pressure is represented by $m^{\text{th}}$-order polynomials that may be either continuous or discontinuous (thus the notation \PN[n]{m} and \PNDG[n]{m}, respectively) across elements. Mass balance (continuity) equations are solved in CV space and a Petrov-Galerkin FEM is used to obtain high-order fluxes on CV boundaries, which are limited to yield bounded fields. Simulations performed for this work were conducted using two types of elements: \PN[1]{2} and \PN[1]{1}. %Saturation, and all saturation-dependent material properties such as relative permeability and capillary pressure, are represented in CV space. %Hence, each CV (and FE) is associated with a unique set of petrophysical properties, in contrast to most previous methods in which CVs span multiple values of petrophysical properties described FE-wise.

@@

%\begin{equation}
%\sum  S_{p}-1
%\label{saturation_constrain}
%\end{equation}

\noindent Equation~\ref{eqn:saturation_eqn} is discretised in time and space using the $\theta$-method. Summing over all phases yields the global mass balance equation,
\begin{eqnarray}
 && \sum_{\alpha=1}^{N_{p}} \left\{\int\limits_{\Omega_{CVi}} \; M_{i} \; \frac{\phi\left(S_{\alpha i}^{n+1}-S_{\alpha i}^{n}\right)}{\Delta t} dV\right.  + \nonumber\\
 &&  \oint_{\Gamma_{CVi}} \left[\; \theta^{n+1/2}\; {\mathbf n}\cdot {\mathbf u}_{\alpha}^{n+1}S_{\alpha}^{n+1} \; + \; \left(1-\theta^{n+1/2}\right) \; {\mathbf u}_{\alpha}^{n}S_{\alpha}^{n}\right] d\Gamma \;- \nonumber\\
 &&  \left.\int_{\Omega_{CVi}} M_{i} \; \mathcal{S}_{cty,\alpha}^{n+\theta} \; dV\;\right\} =0,
\label{global_mass_balance}
\end{eqnarray}
where $\Omega_{CVi}$ and $\Gamma_{CVi}$ are the volume and boundary of CV i respectively, $M_{i}$ are CV basis functions, \textbf{n} is the outward pointing unit normal vector to the surface of $CV_{i}$ and $n$ is the current time level. $\theta$ varies smoothly between $0.5$ (corresponding to Crank-Nicolson method) and $1$ (corresponding to backward-Euler scheme) to avoid the introduction of spurious oscillations for large grid Courant numbers. In addition to the numerical schemes above the discretised global mass and force balance equations are solved using a multigrid-like approach \citep[see][]{pavlidis2016}. This numerical formulation is fully described by \citet{gomes_2017}\citep[see also][]{salinas2015,adam_2016}.

@@

\medskip
As mentioned above viscous flow instabilities can be found across many research areas, disciplines and scales, from chemical and medical separation systems to large scale geological reservoir fields. \citet{muskat_1934} investigated the flow of a fluid injected between two infinite parallel plates and the impact of capillary number ($N_{c}$),

\begin{equation}
N_{c} = \frac{\mu U}{\gamma},
\label{eq:capillary_number}
\end{equation}

\noindent In Eqn. \ref{eq:capillary_number}, $\mu$ is the viscosity of the fuid that initially fills the Hele-Shaw cell, $\gamma$ is the surface tension and $U$ is a velocity characteristic of the moving interface.

\medskip
In the Hele-Shaw cell, a fluid is introduced between two parallel flat plates separated by an infinitesimally small gap. This experimental apparatus allows the predicted instability to be verified qualitatively and simplifies the problem of flow in both porous and non-porous media.   
%The governing equation of Hele-Shaw flows is identical to that of the inviscid potential flow and to the flow of fluid through a porous medium (Darcy's law) since the Hele-Shaw apparatus may be used for both porous and non-porous media.  
More recently, \citet{howison_2000} and \citet{praud_2005} gave a comprehensive description of the mathematical formulation of immiscible two-phase flows in Hele-Shaw cell (also know as Saffman $\&$ Taylor problem). For a Hele-Shaw cell of a given size the developing flow only depends on the capillary number. Thus if \textit{$N_{c}$} is too high \citet{saffman_1959}, \citet{homsy_1987}, \citet{tabeling_1987} and others agreed that the flow develops into a single steady-state finger which moves through the cell with constant velocity $U$.

@@

\medskip 
In heterogeneous domains, such instability may also be triggered by permeability differences across regions as demonstrated in Section \ref{section:results_homo_hete}. The higher the velocity of the low viscosity fluid, the less wide (tip-splitting behaviour) the finger is. Pressure differences acting on the interface produces a net pressure force,

\begin{equation} 
\Delta P= - \gamma \; \nabla \; \cdot \; \hat{n}. 
\label{eq:pressure_dif} 
\end{equation}

@@

\medskip
As demonstrated by \citet{habermann_1960}, \textit{mobility ratio} (MR) is a key-parameter to assess fluid displacement and is defined as the ratio of mobility of the displacing (fluid $i$) to that of the displaced fluid ($j$),
\begin{equation}
 \text{MR} \; = \frac{\mathcal{K}_{ri} \; \mu_{j}}{\mathcal{K}_{rj} \; \mu_{i}}. 
\label{eq:MR}
\end{equation}

\noindent MR is a function not only of fluids' viscosity and saturation, but also of the parameterised relative permeability, $\mathcal{K}_{r\alpha}$, which is often expressed as a function of the residual and maximum phase saturation prescribed in the pore rock matrix. In the simulations conducted for this work, the modified Brooks-Corey model was used \citep{alpak_1999,brooks_1964},

\begin{eqnarray}
  \mathcal{K}_{rw}\left(S_{w}\right) &=& \mathcal{K}^{\circ}_{rw}\left[\frc{S_{w}-S_{w,irr}}{1-S_{w,irr}-S_{nw,r}}\right]^{n_{w}}, \label{Eqn:CoreyBrooks1}\\
  \mathcal{K}_{rnw}\left(S_{w}\right) &=& \mathcal{K}^{\circ}_{rnw}\left[\frc{S_{nw}-S_{nw,r}}{1-S_{w,irr}-S_{nw,r}}\right]^{n_{nw}}, \label{Eqn:CoreyBrooks2}
\end{eqnarray}

@@

\medskip
Here, 2D simulations were conducted in a 5$\times$5 cm domain (Fig.~\ref{fig:homoheleshaw_VN3}a) fully saturated with a fluid, \ie {\it in situ} hydrocarbons. Wetting phase fluid is driven from the bottom left-hand corner of the domain with velocity of 1 cm.s$^{-1}$ (magnitude). Newmann (\ie no flux) boundary conditions were imposed to all borders of the domain except at the top right-hand corner (named as sink), and a pressure gradient between source and sink regions was initially prescribed to the system.
